Subset coercions in \Coq

### 2007-01-23 papers/article/main.tex
Update proof
Correct typeos and some remarks

\documentclass{llncs}
\usepackage{abbrevs}
\usepackage{array}
\usepackage{alltt}
\usepackage{color}
\def\text{\mbox}
\pagestyle{plain}
\usepackage{ifthen}
\usepackage{url}
\usepackage{varioref}

@@

\section{Introduction}
There are many means to program in the \Coq{} environment \cite{CoqArt}. One
can write programs as in \ML{} and prove properties about them
separately, loosing the possibility of using dependent types in
specifications, or give a rich type expressing them as a goal and use the
proof tactics to solve it, producing a corresponding program by the
\CurryHoward isomorphism but having much less control on its algorithmic
essence. It is however difficult to mix the two methods
(writing code and proving interactively) using a rich specification. 
Indeed, when using simple terms and types (\ML{}) or even complex terms and
types (\Coq{}), we can have decidable type-checking. However, when using
simple terms to represent inhabitants of complex types, we immediately get
undecidability of type-checking, as the terms do not give enough
information. Consider for example the function \id{tl} which returns
the tail of a non-empty list. In \ML{}:
\begin{code}
  let tl = function hd :: tl -> tl | [] -> assert(false)
\end{code}

@@

Instead of trying to find a general method for synthetizing proofs from
programs in the \CICfull, we have integrated a method which permits to
link strong specifications and purely algorithmical code. This method,
known as \ps{} in the \PVS{} system, has been used with great success and
fits naturally with how we write specifications using the subset type \cite{progmartinlof}.
The main contribution of this paper is to show how the \ps{} method 
can be adapted in a proof assistant with proof terms, which formally 
justifies the extension in the first place.

@@

\section{\Russell{}}
\label{sec:Russell}
The name of our language is an homage to the mathematician Bertrand Russell who
discovered the famous paradox of set theory with the unrestricted
comprehension axiom. In this theory, it is possible to construct the set
$X = \{x `| x `; x\}$ whose definition is circular. Clearly, if $X `: X$
then $X `; X$ and if $X `; X$ then $X `: X$, hence we have an
inconsistency. \Russell{} was one of the pioneers of type theory when he
devised a set theory with a restricted comprehension axiom that
permitted to create subsets only from already defined subsets,
forbidding the definition of $X$. % \cite{whitehead.russell:principia}.

In \Coq, the distinction between informative and propositional parts of
a term is formalised by the \Prop/\Set{} sorts. In \Russell, we have
special support for propositions appearing in subset types. By
delimiting the use of propositions, we can separate code from proof.

Informally, this idea is already present in mathematics. When you have
an element of subset $\{ x `: S `| P \}$, you can freely forget about the property $P$
and use any operation which is defined on $S$. Conversely, when you
want to use an operation defined on a subset, say $f : \{ x : `N `| P \} "->"
X$, you usually prove \emph{first} that you apply it only to elements 
having the desired property. For example $(f~2)$ is a correct application
only if $(P~2)$ is provable.
In the context of formal development of programs, such a workflow is not
entirely satisfactory because it forces one to create objects and prove
properties about them \emph{at the same time}. In \Coq{}, we are forced
to apply $f$ to an object $(\elt{\nat}{P}{2}{p})$ where $p$ is a proof of
$(P~2)$ (Figure \ref{fig:cci-subset} presents the definition of the subset type in \Coq).
We would like to be able to prove that our usage of partial functions 
is correct only after the program is written. Subsets are particularly well-suited in this respect
because they separate the objects we want to manipulate and their associated
properties.

@@

\subsection{From \ps{} to subset equivalence}
The \ps{} mechanism \cite{Shankar&Owre:WADT99} %Rushby96:TSE
is an extension of the \PVS type system which internalises this idea.
Concretely, this means that the following rules are derivable in \PVS:
\begin{center}
  \AXC{$`G \vdash t : \{ x : T `| P[x] \}$}
  \UIC{$`G \vdash t : T$}
  \DP
  \AXC{$`G \vdash t : T$}
  \AXC{$\vdash_{`G} P[t]$}
  \BIC{$`G \vdash t : \{ x : T `| P[x] \}$}
  \DP
\end{center}

The first one formalises the fact that an object of a subset based on
$T$ is an object of type $T$. The second one permits using an object of
type $T$ as an object of type $\{ x : T `| P[x] \}$, but it generates a
\emph{type-checking condition} $\vdash_{`G} P[t]$ which will need to be
discharged later. Effectively, the typing algorithm of \PVS{} collects
the conditions that must be satisfied for the term to be accepted as a
valid definition. However, the acceptance criteria can be rather large in
\PVS{}. By design, when proving in \PVS{}, the trusted code base (TCB
hereafter) is the entire system, not only the typing system but also the
various decision procedures and tactics used to build proofs
automatically or interactively. It is nonetheless a widely used proof
assistant and the predicate subtyping feature has apparently helped to
build a consequent library of certified code. On the other hand, \Coq
has a small TCB and greater expressiveness but less automation and
methodology to build certified programs. We capitalise on the \PVS
success to make \Coq more usable for this kind of tasks, and as we will see,
it will have other benefits.

@@

Following the presentation as a Pure Type System, the set of sorts
$\PTSsorts$ is defined as $\setproptype$. As usual, we let $s, s_i$ for
$i `: `N$ range over sorts. The axioms are $\PTSaxioms = \{ (\Set,
\Type), (\Prop, \Type) \}$ and the set of rules $\PTSrules$ is defined
by the functional relation $`A s_1 s_2, (s_1, s_2, s_2) `: \PTSrules$. We allow products $\Pi x :
A, B$ where $A : \Prop$ and $B : \Set$ but the user is encouraged to
encode them as $\Pi x : \{ \_ : unit `| A \}, B$ to benefit from
the subtyping mechanism which we will define later.

The reduction rules of this system are the standard $\beta$ and $\pi$
for projections. We denote by $\downarrow{x}$ the complete normal form
of $x$ and by $\hnf{x}$ its head normal form. We use a standard
judgemental equality $`G \type T \eqbp U : s$.

We allow to form dependent sums $\Sigma x : U.V$ (\irule{Sigma},
\irule{Sum-Dep}, \irule{Pi-1}, \irule{Pi-2} rules) only when $U, V :
\Set$ or $U, V : \Prop$. The first sum represents dependent pairs,
useful when defining functions returning a tuple, the latter represent
dependent pairs of propositions, which is most often used as conjunction
of propositions. The dependent pair with $U : \Set$ and $V : \Prop$ is
the subset type which we distinguish (rule \irule{Subset}). The last
possible pair $U : \Prop, V : \Set$ is forbidden because it corresponds
to a pair where the last component, which is informative, may depend on
a particular proof of the proposition $U$. This is quite contrary to the
mantra ``Computations do not depend on proofs'' which governs our
programming language. If there is no dependence then $U$ and $V$ can be
swapped.

@@

\subsubsection{Properties}
We have proved the metatheory for this system in \Coq{}
(\url{http://www.lri.fr/~sozeau/research/russell/proof.en.html}).
 We have assumed that it is strongly
normalising (SN) but we have proved subject reduction (SR) for it: 
if $`G \typed t : T$ and $t "->"_{\beta\pi} t'$ then $`G \typed t' :
T$ (our proof of SR does not depend on SN).
Gang Chen \cite{ChenPhD} has studied various type systems from the
$\lambda$-cube extended by subtyping or coercive subtyping
\cite{DBLP:conf/csl/Luo96}, including the Calculus of Constructions, and
proved such results as SN and SR for them. It requires a very careful
analysis of the system to avoid cycles in the proof due to the presence
of dependent types and conversion. However we preferred to adapt the method of
Robin Adams \cite{adams:PTSEQ} to prove SR, because Gang Chen's method
seemed very tied to the peculiarities of the system he studied. 
Strong normalisation could also be proved, perhaps more easily,
using the generic proof method developed by Geuver \cite{geuvers95}, by finding
a proper interpretation of \Russell terms into the Calculus of
Constructions.

@@

\end{paragraph}
\begin{paragraph}{}
  The next thing to do is to construct a decidable judgement for
  equivalence given two types. We denote it by $`G \typea T
  \suba U : s$ (figure \ref{fig:subtyping-algo-rules}).
  In our case, we have to check that conversion is decidable, which is the
  case as we can replace judgemental equality with the usual untyped
  convertibility $\eqbp$ (we proved $`G \type T \eqbp U : s
  "<=>" T \eqbp U `^ `G \type T, U : s$).
  We also need to eliminate the transitivity and symmetry rules which are not
  syntax-directed. That is the purpose of the following theorem:
\begin{theorem}[Admissibility of transitivity and symmetry]
  \begin{enumerate}
  \item If $`G \typea T \suba U : s$ and $`G \typea U \suba V : s$, then $`G \typea T
    \suba V : s$.
  \item If $`G \typea T \suba U : s$ then $`G \typea U \suba T : s$.
  \end{enumerate}
\end{theorem}
\begin{proof}
  \begin{enumerate}
  \item By induction on the sum of the depths of the two derivations.
  \item By induction, using a thinning lemma.
  \end{enumerate}
\end{proof}
\end{paragraph}

@@

We are ready to build the interpretation, which should have the following property:
\begin{proposition}[Interpretation correctness]
  \label{interp-correct}
  If $`G \typea t : T$ then $\ipG{`G} \typeq \ip{t}{`G} : \ip{T}{`G}$.
\end{proposition}

The proof of this proposition is the main technical contribution of this
paper. It also represented the main difficulty in our work. 
The remainder of this section is organised as follows: first we 
define the explicit coercion derivation algorithm which will permit to
put proof obligations in the terms, then we define the
interpretation of terms which is mutually recursive with the previous
algorithm. Finally we present a proof sketch of the aforementioned
proposition.

@@

Coercions are formalised as multi-holes evaluation contexts, with
$\ctxdot$ the denotation of a hole. We define the instantiation $C[x]$ of a context $C$
by a term $x$ as simultaneous substitution of $x$ for every $\ctxdot$ in $C$.
Hence if $C$ is an evaluation context and $x$ a term, $C[x]$ is a term
(i.e., it cannot contain $\ctxdot$).
The holes will denote the object to which the coercion is applied.
Other
presentations are possible, notably the ``\LMS{}'' style of \cite{longo95logic} which
was adopted by Gang Chen in his treatment of coercive subtyping for the
Calculus of Constructions \cite{conf/popl/Chen03}, and the presentation 
of Amokrane Sa\"ibi \cite{saibi97inheritance}. The \LMS{} style consists
in defining a judgement $x : A \type y : B$ which reads: $A$ is
coercible to $B$ and the coercion is $\lambda x : A.y$. This
presentation is very elegant and adapted for checking coherence 
of coercions, which we do not need here as we can get unicity
of coercions. Moreover, it did not seem readily usable in
our setting, where coercions may contain metavariables. Similarly, Amokrane Sa\"ibi defines
a coercion derivation algorithm which respects some coherence criteria,
roughly, that type-erased coercions $\beta\eta$-reduce to the
identity. Again, this was not adapted to our setting, so we invented this
new presentation.

@@

Now for the original part, the \irule{SubSub} rule simply inserts a
projection before applying the inductively defined coercion on the
support of the subset. On the other hand, \irule{SubProof} creates an
object of a subset using its sole constructor \sref{elt}. The object
part is given by the coercion to the support of the target subset
while the proof part is promised using a metavariable of type $\ip{P}{`G, x : U}[c/x]$ which
we will turn later into a proof obligation.

@@

Similarly, we must add $\eta$ rules for all introduction terms of our
language if we want to be conservative and have unicity of
coercions. The latter is an important property of our system, as it
corresponds in practice to the determinism of proof-obligations
generation. The complete equational theory of \CICq{} is given in figure
\vref{fig:eqccq}. It includes $\eta$, surjective pairing for dependent
sums and subsets ($\rho$) and proof-irrelevance for the second component
of subset objects.
Benjamin Werner has studied the addition of proof-irrelevance (rule
$\sigma$) in the Calculus of Constructions
\cite{Werner:ProofIrrelevance}. It is not a trivial extension and it has
far-reaching consequences on the model of the calculus
\cite{DBLP:conf/types/MiquelW02}. We direct the reader to these papers for further
information on the subject.
\begin{figure}[ht]
  \vspace{-1em}
  \ccqeqarr
  \vspace{-1em}
  \caption{Equational theory of \CICq}
  \label{fig:eqccq}
  \vspace{-1em}
\end{figure}

@@

We first prove reflexivity, symmetry and transitivity of the coercion
derivation algorithm, which extends previous proofs for the algorithmic
system with properties on
the generated coercions. Then we show substitutivity of the
interpretation: if $`G, x : U, `D \type t : T$ then $\ip{t[u/x]}{`G,
 `D[u/x]} = \ip{t}{`G, x : U, `D}[\ip{u}{`G}/x]$. We extend the notion
of coercion to contexts next and show stability of judgements under
context coercion. It is then possible to prove a lemma about commutativity of
substitutions with the interpretation and coercions. Then we get an
important corollary: if $`G \type u : U$, $\subimpl{`G}{c}{U}{V}{s}$ and
$`G, x : V \type T : s$ then $\ip{T[u/x]}{`G} `= \ip{T}{`G, x :
  V}[c[\ip{u}{`G}]/x]$. Essentially, it shows that coercions inserted
by the interpretation depend only on the context and hence could be
added later using substitution instead. This allows us to show that
equivalence is conserved by interpretation and the following lemma:
\begin{lemma}[Coercion derivation correctness]
  \label{lemma:coercion-correct}
  If $\subimpl{`G}{c}{T}{U}{s}$ then $\ipG{`G} \typec \lambda x :
  \ip{T}{`G}.c[x] : \ip{T}{`G} "->" \ip{U}{`G}$.
\end{lemma}
 Finally we can show that our interpretation is correct (proposition \ref{interp-correct}).

@@

\subsection{Related work}
Other efforts to build certified programs using type theory include
\Cayenne \cite{Augustsson99} and \Epigram
\cite{DBLP:journals/jfp/McBrideM04}. \Cayenne offers dependent
types and general recursion at the expense of a non-terminating type
checker; it can be thought of as a testbed for developing languages with
dependent types but is not so much aimed at building certified programs.
\Epigram ought to become a complete programming language with dependent
types at its heart. Instead of using a phase distinction to separate
coding and proving, \Epigram is based on an interactive, 2-dimensional editing process
where code and proofs are written incrementally to get a complete program.
Using type annotations and with the help of the editor for structuring, 
one is able to write programs with precis e specifications. However it
is not obvious if it could be made usable by programmers because of the
paradigm change and scaling issues. The \DML language \cite{Xi98:PhD} is more akin to
comparison with our solution. It provides a way to use a restricted set
of dependent types in \ML programs, carefully chosen so as to keep
type-checking decidable with the help of an automatic prover. Our method
should subsume this one as it is perfectly possible to integrate automatic
proof tools with \Coq to discharge the generated obligations.

@@

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: t
%%% LaTeX-command: "BIBINPUTS=\"${HOME}/research/demons-biblio:../bib:$BIBINPUTS\" TEXINPUTS=\"..:../style:$TEXINPUTS\" latex"
%%% End:

>>>>

\documentclass{llncs}
\usepackage{abbrevs}
%\usepackage{array}
\usepackage{alltt}
\def\text{\mbox}

%%%% FixMe: remove 
\pagestyle{plain}

\usepackage{ifthen}
\usepackage{url}
\usepackage{varioref}

@@

\section{Introduction}
There are many means to program in the \Coq{} environment \cite{CoqArt}. One
can write programs as in \ML{} and prove properties about them
separately, losing the possibility of using dependent types in
specifications, or give a rich type expressing them as a goal and use the
proof tactics to solve it, producing a corresponding program by the
\CurryHoward isomorphism but having much less control on its algorithmic
essence. It is however difficult to mix the two methods
(writing code and proving interactively) using a rich specification. 
Indeed, when using simple terms and types (\ML{}) or even complex terms and
types (\Coq{}), we can have decidable type-checking. However, when using
simple terms to represent inhabitants of complex types, we immediately get
undecidability of type-checking, as the terms do not give enough
information. Consider for example the function \id{tl} which returns
the tail of a non-empty list. In \ML{}:
\begin{code}
  let tl = function hd :: tl -> tl | [] -> assert(false)
\end{code}

@@

Instead of trying to find a general method for synthesizing proofs from
programs in the \CICfull, we have integrated a method which permits to
link strong specifications and purely algorithmical code. This method,
known as \ps{} in the \PVS{} system, has been used with great success and
fits naturally with how we write specifications using the subset type \cite{progmartinlof}.
The main contribution of this paper is to show how the \ps{} method 
can be adapted in a proof assistant with proof terms, which formally 
justifies the extension in the first place.

@@

\section{\Russell{}}
\label{sec:Russell}
The name of our language is an homage to the mathematician Bertrand Russell who
discovered the famous paradox of set theory with the unrestricted
comprehension axiom. In this theory, it is possible to construct the set
$X = \{x `| x `; x\}$ whose definition is circular. Clearly, if $X `: X$
then $X `; X$ and if $X `; X$ then $X `: X$, hence we have an
inconsistency. Russell was one of the pioneers of type theory when he
devised a set theory with a restricted comprehension axiom that
permitted to create subsets only from already defined subsets,
forbidding the definition of $X$. % \cite{whitehead.russell:principia}.

In \Coq, the distinction between informative and propositional parts of
a term is formalised by the \Set{}/\Prop{} sorts. In \Russell, we have
special support for propositions appearing in subset types. By
delimiting the use of propositions, we can separate code from proof.

Informally, this idea is already present in mathematics. When you have
an element of subset $\{ x `: S `| P \}$, you can freely forget about the property $P$
and use any operation which is defined on $S$. Conversely, when you
want to use an operation defined on a subset, say $f : \{ x : `N `| P \} "->"
X$, you usually prove \emph{first} that you apply it only to elements 
having the desired property. For example $(f~2)$ is a correct application
only if $(P[2/x])$ is provable.
In the context of formal development of programs, such a workflow is not
entirely satisfactory because it forces one to create objects and prove
properties about them \emph{at the same time}. In \Coq{}, we are forced
to apply $f$ to an object $(\elt{\nat}{\lambda x : \nat. P}{2}{p})$ where $p$ is a proof of
$(P[2/x])$ (Figure \ref{fig:cci-subset} presents the definition of the subset type in \Coq).
We would like to be able to prove that our usage of partial functions 
is correct only after the program is written. Subsets are particularly well-suited in this respect
because they separate the objects we want to manipulate and their associated
properties.

@@

\subsection{From \ps{} to subset equivalence}
The \ps{} mechanism \cite{Shankar&Owre:WADT99} %Rushby96:TSE
is an extension of the \PVS type system which internalises this idea.
Concretely, this means that the following rules are derivable in \PVS:
\begin{center}
  \AXC{$`G \vdash t : \{ x : T `| P[x] \}$}
  \UIC{$`G \vdash t : T$}
  \DP
  \AXC{$`G \vdash t : T$}
  \AXC{$`G \vdash P[t]$}
  \BIC{$`G \vdash t : \{ x : T `| P[x] \}$}
  \DP
\end{center}

The first one formalises the fact that an object of a subset based on
$T$ is an object of type $T$. The second one permits using an object of
type $T$ as an object of type $\{ x : T `| P[x] \}$, but it generates a
\emph{type-checking condition} $`G \vdash P[t]$ which will need to be
discharged later. Effectively, the typing algorithm of \PVS{} collects
the conditions that must be satisfied for the term to be accepted as a
valid definition. However, the acceptance criteria can be rather large in
\PVS{}. By design, when proving in \PVS{}, the trusted code base (TCB
hereafter) is the entire system, not only the typing system but also the
various decision procedures and tactics used to build proofs
automatically or interactively. It is nonetheless a widely used proof
assistant and the predicate subtyping feature has apparently helped to
build a consequent library of certified code. On the other hand, \Coq
has a small TCB and greater expressiveness but less automation and
methodology to build certified programs. We capitalise on the \PVS
success to make \Coq more usable for this kind of tasks, and as we will see,
it will have other benefits.

@@

Following the presentation as a Pure Type System, the set of sorts
$\PTSsorts$ is defined as $\setproptype$. As usual, we let $s, s_i$ for
$i `: `N$ range over sorts. The axioms are $\PTSaxioms = \{ (\Set,
\Type), (\Prop, \Type) \}$ and the set of rules $\PTSrules$ is defined
by the functional relation $`A s_1 s_2, (s_1, s_2, s_2) `: \PTSrules$. We allow products $\Pi x :
A. B$ where $A : \Prop$ and $B : \Set$ but the user is encouraged to
encode them as $\Pi x : \{ \_ : unit `| A \}. B$ to benefit from
the subtyping mechanism which we will define later.

The reduction rules of this system are the standard $\beta$ and $\pi$
for projections. We denote by $\downarrow{x}$ the complete normal form
of $x$ and by $\hnf{x}$ its weak head normal form. We use a standard
judgemental equality $`G \type T \eqbp U : s$.

We allow to form dependent sums $\Sigma x : U.V$ (\irule{Sigma},
\irule{Pair}, \irule{Pi-1}, \irule{Pi-2} rules) only when $U, V :
\Set$ or $U, V : \Prop$. The first sum represents dependent pairs,
useful when defining functions returning a tuple, the latter represent
dependent pairs of propositions, which is most often used as conjunction
of propositions. The dependent pair with $U : \Set$ and $V : \Prop$ is
the subset type which we distinguish (rule \irule{Subset}). The last
possible pair $U : \Prop, V : \Set$ is forbidden because it corresponds
to a pair where the last component, which is informative, may depend on
a particular proof of the proposition $U$. This is quite contrary to the
mantra ``Computations do not depend on proofs'' which governs our
programming language. If there is no dependence then the components can be
swapped.

@@

\subsubsection{Properties}
We have proved some of the metatheory for this system in \Coq{}
\cite{sozeau:Coq/Russell/meta-web}.
We have assumed that it is strongly
normalising (SN) but we have proved subject reduction (SR) for it: 
if $`G \typed t : T$ and $t "->"_{\beta\pi} t'$ then $`G \typed t' :
T$ (our proof of SR does not depend on SN).
Gang Chen \cite{ChenPhD} has studied various type systems from the
$\lambda$-cube extended by subtyping or coercive subtyping
\cite{DBLP:conf/csl/Luo96}, including the Calculus of Constructions, and
proved such results as SN and SR for them. It requires a very careful
analysis of the system to avoid cycles in the proof due to the presence
of dependent types and conversion. However we preferred to adapt the method of
Robin Adams \cite{adams:PTSEQ} to prove SR, because Gang Chen's method
seemed very tied to the peculiarities of the system he studied. 
Strong normalisation could also be proved, perhaps more easily,
using the generic proof method developed by Geuvers \cite{geuvers95}, by finding
a proper interpretation of \Russell terms into the Calculus of
Constructions.

@@

\end{paragraph}
\begin{paragraph}{}
  The next thing to do is to construct a decidable judgement for
  equivalence given two types. We denote it by $`G \typea T
  \suba U : s$ (figure \ref{fig:subtyping-algo-rules}).
  In our case, we have to check that conversion is decidable, which is the
  case as we can replace judgemental equality with the usual untyped
  convertibility $\eqbp$ (we proved $`G \type T \eqbp U : s
  "<=>" T \eqbp U `^ `G \type T, U : s$ in \Coq). 
  Clearly, we can eliminate symmetry directly, given the fact that
  conversion is symmetric.
  We also need to eliminate the transitivity rule which is not
  syntax-directed. That is the purpose of the following theorem:
\begin{theorem}[Admissibility of transitivity and symmetry]
  \begin{enumerate}
  \item If $`G \typea T \suba U : s$ and $`G \typea U \suba V : s$, then $`G \typea T
    \suba V : s$.
  \item If $`G \typea T \suba U : s$ then $`G \typea U \suba T : s$.
  \end{enumerate}
\end{theorem}
\begin{proof}
  \begin{enumerate}
  \item By induction on the sum of the depths of the two derivations.
  \item By induction, using a thinning lemma.
  \end{enumerate}
\end{proof}
\end{paragraph}

@@

We are ready to build the interpretation, which should have the following property:
\begin{theorem}[Interpretation correctness]
  \label{interp-correct}
  If $`G \typea t : T$ then $\ipG{`G} \typeq \ip{t}{`G} : \ip{T}{`G}$.
\end{theorem}

The proof of this proposition is the main technical contribution of this
paper. It also represented the main difficulty in our work. 
The remainder of this section is organised as follows: first we 
define the explicit coercion derivation algorithm which will permit to
put proof obligations in the terms, then we define the
interpretation of terms which is mutually recursive with the previous
algorithm. Finally we present a proof sketch of the aforementioned
theorem.

@@

Coercions are formalised as multi-holes evaluation contexts, with
$\ctxdot$ the denotation of a hole. We define the instantiation $C[x]$ of a context $C$
by a term $x$ as simultaneous substitution of $x$ for every $\ctxdot$ in $C$.
Hence if $C$ is an evaluation context and $x$ a term, $C[x]$ is a term
(i.e., it cannot contain $\ctxdot$).
The holes will denote the object to which the coercion is applied.
Other
presentations are possible, notably the ``\LMS{}'' style of \cite{longo95logic} which
was adopted by Gang Chen in his treatment of coercive subtyping for the
Calculus of Constructions \cite{conf/popl/Chen03}, and the presentation 
of Amokrane Sa\"ibi \cite{saibi97inheritance}. The \LMS{} style consists
in defining a judgement $x : A \type y : B$ which reads: $A$ is
coercible to $B$ and the coercion is $\lambda x : A.y$. This
presentation is very elegant and adapted for checking coherence 
of coercions, which we do not need here as we can get uniqueness
of coercions. Moreover, it did not seem readily usable in
our setting, where coercions may contain metavariables. Similarly, Amokrane Sa\"ibi defines
a coercion derivation algorithm which respects some coherence criteria,
roughly, that type-erased coercions $\beta\eta$-reduce to the
identity. Again, this was not adapted to our setting, so we invented this
new presentation.

@@

Now for the original part, the \irule{SubSub} rule simply inserts a
projection before applying the inductively defined coercion on the
support of the subset. On the other hand, \irule{SubProof} creates an
object of a subset using its sole constructor \sref{elt} (with type
parameters omitted for brevity). The object
part is given by the coercion to the support of the target subset
while the proof part is promised using a metavariable of type $\ip{P}{`G, x : U}[c/x]$ which
we will turn later into a proof obligation.

@@

Similarly, we must add $\eta$ rules for all introduction terms of our
language if we want to be conservative and have uniqueness of
coercions. The latter is an important property of our system, as it
corresponds in practice to the determinism of proof-obligations
generation. The complete equational theory of \CICq{} is given in figure
\vref{fig:eqccq}. It includes $\eta$, surjective pairing for dependent
sums and subsets ($\rho$) and proof-irrelevance for the second component
of subset objects.
Benjamin Werner has studied the addition of proof-irrelevance (rule
$\sigma$) in the Calculus of Constructions
\cite{Werner:ProofIrrelevance}. It is not a trivial extension and it has
far-reaching consequences on the model of the calculus
\cite{DBLP:conf/types/MiquelW02}. We direct the reader to these papers for further
information on the subject.
\begin{figure}[ht]
  \vspace{-1em}
  \ccqeqarr
  \vspace{-1em}
  \caption{Equational theory of \CICq}
  \label{fig:eqccq}
  \vspace{-1em}
\end{figure}

@@

We first prove reflexivity, symmetry and transitivity of the coercion
derivation algorithm, which extends previous proofs for the algorithmic
system with properties on
the generated coercions. Then we show substitutivity of the
interpretation: if $`G, x : U, `D \type t : T$ then $\ip{t[u/x]}{`G,
 `D[u/x]} = \ip{t}{`G, x : U, `D}[\ip{u}{`G}/x]$. We extend the notion
of coercion to contexts next and show stability of judgements under
context coercion. It is then possible to prove a lemma about commutativity of
substitutions with the interpretation and coercions. Then we get an
important corollary: if $`G \type u : U$, $\subimpl{`G}{c}{U}{V}{s}$ and
$`G, x : V \type T : s$ then $\ip{T[u/x]}{`G} `= \ip{T}{`G, x :
  V}[c[\ip{u}{`G}]/x]$. Essentially, it shows that coercions inserted
by the interpretation depend only on the context and hence could be
added later using substitution instead. This allows us to show that
equivalence is conserved by interpretation and the following lemma:
\begin{lemma}[Coercion derivation correctness]
  \label{lemma:coercion-correct}
  If $\subimpl{`G}{c}{T}{U}{s}$ then $\ipG{`G} \typec \lambda x :
  \ip{T}{`G}.c[x] : \ip{T}{`G} "->" \ip{U}{`G}$.
\end{lemma}
Finally we can show that our interpretation is correct (theorem \ref{interp-correct}).

@@

\subsection{Related work}
Other efforts to build certified programs using type theory include
\Cayenne \cite{Augustsson99} and \Epigram
\cite{DBLP:journals/jfp/McBrideM04}. \Cayenne offers dependent
types and general recursion at the expense of a non-terminating type
checker; it can be thought of as a testbed for developing languages with
dependent types but is not so much aimed at building certified programs.
\Epigram ought to become a complete programming language with dependent
types at its heart. Instead of using a phase distinction to separate
coding and proving, \Epigram is based on an interactive, 2-dimensional editing process
where code and proofs are written incrementally to get a complete program.
Using type annotations and with the help of the editor for structuring, 
one is able to write programs with precise specifications. However it
is not obvious if it could be made usable by programmers because of the
paradigm change and scaling issues. The \DML language \cite{Xi98:PhD} is more akin to
comparison with our solution. It provides a way to use a restricted set
of dependent types in \ML programs, carefully chosen so as to keep
type-checking decidable with the help of an automatic prover. Our method
should subsume this one as it is perfectly possible to integrate automatic
proof tools with \Coq to discharge the generated obligations.

@@

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: t
%%% LaTeX-command:
%%% "BIBINPUTS=\"${HOME}/research/demons-biblio:../bib:$BIBINPUTS\"
%%% TEXINPUTS=\"${HOME}/.latex:..:../style:$TEXINPUTS\" latex"
%%% End:
